---
title: "TCIA Data Exploration"
author: "Grant Carr"
date: "`r Sys.Date()`"
output: pdf_document
editor_options: 
  markdown: 
    wrap: 72
---

## Purpose

Exploratory data analysis is always the first step in a research
project. The purpose is to explore whether there may be evidence of
patterns in the data and to make sure you have an understanding of what
variables are relevant prior to pursuing a research question.

This markdown is meant as an introduction to exploratory data analysis
in R. It is not comprehensive. There are relevant clinical variables and
imaging features that may be interesting to you, but you are not
explicitly asked to explore in the tasks below. You should explore more
outcomes and imaging features beyond those listed in the tasks in order
to gain a better understanding of the data.

## R Setup

Load packages that are necessary for running all code below.

```{r}
#' example code for installing packages for first time:
#' install.packages("tidyverse")

library(tidyverse)
library(ggplot2)
library(survival)
library(ggfortify)
library(devtools)
#' install_github("jokergoo/ComplexHeatmap")
library(ComplexHeatmap)
```

## Reading Clinical Data

```{r}
clinData <- read.csv(
  "/Users/lucymalmud/Desktop/BDSI/cancer-ds/tcia-git/data/clinicalData_clean.csv"
)
clinData <- clinData[,-1] # remove first column of row numbers since we have
# patient ID
clinData %>% head
```

The '\<-' is the assignment operator in R. clinData \<- read.csv(...)
assigns the output of read.csv() into a variable/object called clinData.
Keyboard shortcut on a Mac is option, -. You can also use '='. '\<-' is
used to differentiate setting function arguments versus assigning data
to a variable.

## Reading Image Features

```{r}
imFeatures <- read.csv(
  "/Users/lucymalmud/Desktop/BDSI/cancer-ds/tcia-git/data/imagingFeatures.csv"
)
imFeatures <- imFeatures[,-1]
```

In general, the featureCitations.docx file describes the features and
provides citations. The features are grouped into the categories listed
below.

Breast and fibroglandular tissue (FGT) volume features: volume and
density of breast area and fibroglandular area.

Tumor size and morphology: regularity/roundness vs irregularity of tumor
shape and size.

FGT enhancement: measure how much FGT is enhanced when we add contrast.
This is referred to as BPE, background parenchymal enhancement, which
may confound tumor enhancement/identification.

Tumor enhancement: measure how much tumor is enhanced when we add
contrast

Combining tumor and FGT enhancement: measure how tumor and FGT are
enhanced when adding contrast

FGT enhancement texture: describe the enhancement due to contrast.
Looking at local patterns, does enhancement look gritty or smooth?

Tumor enhancement texture: same as FGT but for tumor area.

Tumor Enhancement Spatial Heterogeneity: measure similarity between
tumor subregions of the tumor.

FGT enhancement variation: global variation of contrast enhancement of
FGT

Tumor enhancement variation: global variation of contrast enhancement of
tumor

## Task 1: What is in the data?

How many patients are in the dataset?

How many missing values are there among image features?

How many missing values are there among clinical features?

What types of variables are in the clinical dataset? Are they
quantitative? Are they qualitative/categorical? If they are categorical,
how many categories are there? Pay attention especially to variables
such as time to event, receptor positivity, and tumor characteristics
since these are potentially relevant to research questions.

## Task 2: Visualizing Data

Tabulate some discrete clinical variables and plot some continuous
clinical variables. Take note of any abnormalities such as missingness,
low data in a category, or significant skew in continuous variables.

Pick a subset of image features you are interested in. It can be
arbitrary or defined by the groups in featureDocumentation.docx. Explore
whether there is an association between the image feature(s) and any
clinical variable(s) that you are interested in. You can use
scatterplots or any visualization tool you see fit.

```{r}
clinData %>% select('Date.of.Birth..Days.', 'Days.to.death..from.the.date.of.diagnosis.') %>% 
  filter(Days.to.death..from.the.date.of.diagnosis. != 'NP') %>% 
  ggplot() +
  aes(x = Date.of.Birth..Days.,
      y = Days.to.death..from.the.date.of.diagnosis.) +
  geom_point()
```

It does not look like there is a correlation between days to death and
birth date.

Next, find a function from an R package that can create a "heat map" and
see if the features cluster together. Overlay your clustering with a
clinical feature. Does the clustering of image features seem to
correlate with the clinical feature?

```{r}
clinData %>% select('Mol.Subtype', 'Echogenicity') %>% 
  filter(Echogenicity != 'NC' & Echogenicity != 'NA') %>% 
  mutate(Echogenicity = case_when(Echogenicity %in% c("0", "0 (0)", "0 [0]", "[0]") ~ "0",
                                  Echogenicity %in% c("3", "[3]") ~ "3",
                                  Echogenicity == "[2]" ~ "2",
                                  Echogenicity == "2 [0]" ~ "2, 0",
                                  Echogenicity == "3[5]" ~ "3, 5",
                                  .default = Echogenicity)) %>% 
  ggplot() +
  aes(x = Mol.Subtype,
      y = Echogenicity) +
  geom_point()
```

It does not seem like there is any correlation between echogenicity and
molecular subtype.

Heat maps can be a useful tool for finding broad patterns in data. In
the above example, there may be some clustering of textural features,
but the clustering does not correspond to any clinical outcomes.
Clustering does not provide rigorous statistical evidence of
associations or lack thereof, but it can be a useful exploratory step.

## Task 3: Preliminary Modeling

Consider HER2 positivity as a binary outcome. Fit a logistic regression
model using the covariates used above. Which variables are statistically
significant based on a significance level of 0.05?

## Task 4: Survival Curves

Fit a Kaplan-Meier survival curve to the time to death and time to
recurrence data. For each time to event outcome, fit a Kaplan-Meier
curve split by molecular subtype. Which subtypes have the most favorable
prognosis? Least favorable?
