---
title: "TCIA Data Exploration"
author: "Grant Carr"
date: "`r Sys.Date()`"
output: pdf_document
editor_options: 
  markdown: 
    wrap: 72
---

## Purpose

Exploratory data analysis is always the first step in a research
project. The purpose is to explore whether there may be evidence of
patterns in the data and to make sure you have an understanding of what
variables are relevant prior to pursuing a research question.

This markdown is meant as an introduction to exploratory data analysis
in R. It is not comprehensive. There are relevant clinical variables and
imaging features that may be interesting to you, but you are not
explicitly asked to explore in the tasks below. You should explore more
outcomes and imaging features beyond those listed in the tasks in order
to gain a better understanding of the data.

## R Setup

Load packages that are necessary for running all code below.

```{r}
#' example code for installing packages for first time:
#' install.packages("tidyverse")

library(tidyverse)
library(ggplot2)
library(survival)
library(ggfortify)
library(devtools)
#' install_github("jokergoo/ComplexHeatmap")
library(ComplexHeatmap)
```

## Reading Clinical Data

```{r}
clinData <- read.csv(
  "/Users/grantcarr/Documents/Michigan/BDSI2025/Data/clinicalData_clean.csv"
)
clinData <- clinData[,-1] # remove first column of row numbers since we have
# patient ID

# Create a histogram showing which patients had surgery
clinData %>% select("Surgery") %>% ggplot() +
  aes(x=Surgery) +
  geom_histogram()
```

The '\<-' is the assignment operator in R. clinData \<- read.csv(...)
assigns the output of read.csv() into a variable/object called clinData.
Keyboard shortcut on a Mac is option, -. You can also use '='. '\<-' is
used to differentiate setting function arguments versus assigning data
to a variable.

## Reading Image Features

```{r}
imFeatures <- read.csv(
  "/Users/grantcarr/Documents/Michigan/BDSI2025/Data/imagingFeatures.csv"
)
imFeatures <- imFeatures[,-1]
```

In general, the featureCitations.docx file describes the features and
provides citations. The features are grouped into the categories listed
below.

Breast and fibroglandular tissue (FGT) volume features: volume and
density of breast area and fibroglandular area.

Tumor size and morphology: regularity/roundness vs irregularity of tumor
shape and size.

FGT enhancement: measure how much FGT is enhanced when we add contrast.
This is referred to as BPE, background parenchymal enhancement, which
may confound tumor enhancement/identification.

Tumor enhancement: measure how much tumor is enhanced when we add
contrast

Combining tumor and FGT enhancement: measure how tumor and FGT are
enhanced when adding contrast

FGT enhancement texture: describe the enhancement due to contrast.
Looking at local patterns, does enhancement look gritty or smooth?

Tumor enhancement texture: same as FGT but for tumor area.

Tumor Enhancement Spatial Heterogeneity: measure similarity between
tumor subregions of the tumor.

FGT enhancement variation: global variation of contrast enhancement of
FGT

Tumor enhancement variation: global variation of contrast enhancement of
tumor

## Task 1: What is in the data?

How many patients are in the dataset?

```{r}
nrow(clinData)
```

How many missing values are there among image features?

```{r}
sum(is.na(imFeatures))

missingness <- apply(
  imFeatures, 2, function(x){
    sum(is.na(x))
  }
)
sum(missingness == 0) # there are 490 variables with no missingness
sum(missingness != 0) # the remaining 40 variables have some missing values
missingness[missingness != 0]
```

How many missing values are there among clinical features?

```{r}
sum(is.na(clinData))

missingness <- apply(
  clinData, 2, function(x){
    sum(is.na(x))
  }
)
sum(missingness == 0) # there are 45 variables with no missingness
sum(missingness != 0) # the remaining 52 variables have some missing values
missingness[missingness != 0]
```

Some clinical variables have minimal missing values, such as "Contrast
Agent" (5). Others have so many missing values that the variable is
statistically useless, such as "Contrast Bolus Volume (mL)" (653).

What types of variables are in the clinical dataset? Are they
quantitative? Are they qualitative/categorical? If they are categorical,
how many categories are there? Pay attention especially to variables
such as time to event, receptor positivity, and tumor characteristics
since these are potentially relevant to research questions.

```{r}
summary(clinData[,"Days.to.MRI..From.the.Date.of.Diagnosis."])
table(clinData$Manufacturer, useNA = "if")
table(clinData$ER, useNA = "if")
table(clinData$PR, useNA = "if")
table(clinData$HER2, useNA = "if")
table(clinData$Mol.Subtype, useNA = "if")


summary(-1*as.numeric(clinData[,"Date.of.Birth..Days."])/365)
# 21.75 - 89.49 years old, median 52.25 years old
sum(is.na(as.numeric(clinData[,"Date.of.Birth..Days."])))

clinData %>% select(Menopause..at.diagnosis.:Mol.Subtype) %>% sapply(table, useNA = "if")
clinData %>% select(Staging.Tumor.Size....T.:Tumor.Location) %>% sapply(table, useNA = "if")
```

## Task 2: Visualizing Data

Tabulate some discrete clinical variables and plot some continuous
clinical variables. Take note of any abnormalities such as missingness,
low data in a category, or significant skew in continuous variables.

```{r}
apply(clinData, 2, function(x){
  sum(is.na(x)) + sum(x %in% c("NA", "NC", "NP"))
}) %>% table()

sum(is.na(clinData$Days.to.MRI..From.the.Date.of.Diagnosis.))
hist(clinData$Days.to.MRI..From.the.Date.of.Diagnosis.)
sum(clinData$Days.to.MRI..From.the.Date.of.Diagnosis. > 0) # 808
sum(clinData$Days.to.MRI..From.the.Date.of.Diagnosis. == 0) # 9
sum(clinData$Days.to.MRI..From.the.Date.of.Diagnosis. < 0) # 105

#' everyone gets MRI at different times. knowledge of when other clinical 
#' data is collected may be important in understanding whether
#' image features can reflect relevant information

clinData <- clinData %>% mutate(
  ageAtDiagnosis = -1*as.numeric(Date.of.Birth..Days.)/365
)
hist(clinData$ageAtDiagnosis)
summary(clinData[,"ageAtDiagnosis"])

clinData <- clinData %>% mutate(
  raceEth = case_when(
    Race.and.Ethnicity == 0 ~ NA,
    Race.and.Ethnicity == 1 ~ "White",
    Race.and.Ethnicity == 2 ~ "Black",
    Race.and.Ethnicity == 3 ~ "Asian",
    Race.and.Ethnicity == 4 ~ "Native",
    Race.and.Ethnicity == 5 ~ "Hispanic",
    Race.and.Ethnicity == 6 ~ "Multi",
    Race.and.Ethnicity == 7 ~ "Hawaiian",
    Race.and.Ethnicity == 8 ~ "American Indian",
    .default = "?"
  )
)
table(clinData$raceEth, useNA = "if")

clinData <- clinData %>% 
  mutate(
    Subtype = factor(
      case_when(
        Mol.Subtype == 0 ~ "Luminal",
        Mol.Subtype == 1 ~ "ER/PR+ and HER2+",
        Mol.Subtype == 2 ~ "HER2+",
        .default = "Triple Negative"
      ), levels = c("Luminal", "ER/PR+ and HER2+", "HER2+", "Triple Negative")
    )
  )

unique(clinData[clinData$Subtype == "Luminal", c("ER", "PR", "HER2")])
unique(clinData[clinData$Subtype == "ER/PR+ and HER2+", c("ER", "PR", "HER2")])
unique(clinData[clinData$Subtype == "HER2+", c("ER", "PR", "HER2")])
unique(clinData[clinData$Subtype == "Triple Negative", c("ER", "PR", "HER2")])
#' Luminal: at least one ER+/PR+ but not HER2+
#' ER/PR and HER2: HER2+ and at least one ER+/PR+
#' HER2+: HER2+ only
#' triple negative: ER-/PR-/HER2-
table(clinData["Subtype"])
# dominated by luminal subtype, some sparsity in others

table(clinData[,"Staging.Tumor.Size....T."])
# dominated by stage 1/2, sparsity in stages 3/4
table(clinData[,"Staging.Nodes...Nx.replaced.by..1..N."])
# dominated by stage 0/1, sparsity in stages 2/3
table(clinData[,"Staging.Metastasis...Mx..replaced.by..1..M."])
# dominated by 0 (no mets) and -1 (can't be evaluated)

#' in general, early stage disease

table(clinData["TumorGradeT"], useNA = "if")
table(clinData["TumorGradeN"], useNA = "if")
table(clinData["TumorGradeM"], useNA = "if")
#' very little missingness for tumor grade
#' high Tubule/Nuclear grade, low Mitotic grade

table(clinData["Nottingham.grade"], useNA = "if")
table(clinData["Histologic.type"], useNA = "if")
#' high degree of missingness in these variables

table(clinData[,"Bilateral.Information"])
# lots of missingness, and very few with bilateral disease

table(clinData["Surgery"], useNA = "if")
# nearly all patients had surgery
clinData <- clinData %>% mutate(
  surgeryDays = as.numeric(Days.to.Surgery..from.the.date.of.diagnosis.)
)
sum(is.na(clinData["surgeryDays"])) # 47 missing values
sum(complete.cases(clinData[ , c("surgeryDays", "Surgery")]))
# 875 observations for time to surgery
table(clinData["Definitive.Surgery.Type"], useNA = "if")
# 50/50 BCS vs mastectomy

sum(
  clinData["Neoadjuvant.Radiation.Therapy"] == 1, na.rm = T
) # 22 patients neoadjuvant radiation therapy (given before surgery)
sum(
  clinData["Adjuvant.Radiation.Therapy"] == 1, na.rm = T
) # 614 patients adjuvant radiation therapy (given after surgery)

clinData$dead <- ifelse(
  clinData$Days.to.death..from.the.date.of.diagnosis. == "NP",
  0, 1
) # if days to death is not pertinent, then they did not die
table(clinData["dead"]) # 62 deaths
clinData$survDays <- ifelse(
  clinData$dead == 1, as.numeric(clinData$Days.to.death..from.the.date.of.diagnosis.),
  pmax(
    as.numeric(clinData$Days.to.last.distant.recurrence.free.assemssment.from.the.date.of.diagnosis.),
    as.numeric(clinData$Days.to.last.local.recurrence.free.assessment..from.the.date.of.diagnosis.),
    as.numeric(clinData$Age.at.last.contact.in.EMR.f.u.days..from.the.date.of.diagnosis...last.time.patient.known.to.be.alive..unless.age.of.death.is.reported.in.such.case.the.age.of.death), na.rm = T
  )
)
#' if dead, then use days to death
#' if they did not die, then use the last time we know any information
#' about them being alive. pmax = parallel maximum of vectors, pmax of 
#' last local recurrence free assessment, last distant recurrence free 
#' assessment, and last contact in electronic medical record
summary(clinData["survDays"]/365) # IQR 2.8-5.2, median 4 years
hist(unlist(clinData["survDays"]))
sum(complete.cases(clinData[ , c("survDays", "dead")]))
# 922 observations for time to death

clinData$recurrence <- as.numeric(clinData$Recurrence.event.s.)
#' make recurrence a numeric indicator
table(clinData["recurrence"], useNA = "if") # 87 recurrence, 2 missing
clinData$recurDays <- ifelse(
  clinData$recurrence == 0, as.numeric(clinData$survDays),
  ifelse(
    clinData$recurrence == 1,
    pmin(
    as.numeric(clinData$Days.to.distant.recurrence.from.the.date.of.diagnosis.),
    as.numeric(clinData$Days.to.local.recurrence..from.the.date.of.diagnosis.),
    na.rm = T
  ),
    NA
  )
  
)
#' if no recurrence, then days to recurrence is just survival days
#' if recurrence, then days to recurrence is minimum of days to local 
#' recurrence and days to distant recurrence
#' if recurrence is missing, then days to recurrence is missing
hist(unlist(clinData["recurDays"]))
sum(complete.cases(clinData[ , c("recurDays", "recurrence")]))
# 920 observations for time to recurrence

table(clinData[ , "Neoadjuvant.Chemotherapy"], useNA = "if")
table(clinData[ , "Adjuvant.Chemotherapy"], useNA = "if")
# many with neoadjuvant or adjuvant chemotherapy

table(clinData[ , "Neoadjuvant.Anti.Her2.Neu.Therapy"], useNA = "if")
table(clinData[ , "Adjuvant.Anti.Her2.Neu.Therapy"], useNA = "if")
# few with neoadjuvant or adjuvant anti-Her2

table(clinData[,"Received.Neoadjuvant.Therapy.or.Not"], useNA = "if")
# overall, most patients did not receive any type of neoadjuvant therapy
```

Pick a subset of image features you are interested in. It can be
arbitrary or defined by the groups in featureDocumentation.docx. Explore
whether there is an association between the image feature(s) and any
clinical variable(s) that you are interested in. You can use
scatterplots or any visualization tool you see fit.

```{r}
#' I am interested in tumor enhancement texture and its relationship
#' to tumor subtype
tumorEnhanceTexture_vars <- colnames(imFeatures)[
  c(189:232, 279:281, 287:308, 313:334, 339:360, 27:48)
]
# manually identify all variables in the tumor enhancement texture group, 
# 135 of them

fullData <- left_join(
  clinData, imFeatures
)

subData <- fullData %>% 
  select(Subtype, all_of(tumorEnhanceTexture_vars))
ggplot(
  subData, 
  aes(x = Subtype, y = X1st_DFT_CoeffMap_Momment_Invariant_1_3D_tumor)
) +
  geom_boxplot()
ggplot(
  subData, 
  aes(x = Subtype, y = SER_map_Autocorrelation_tumor)
) +
  geom_boxplot()
ggplot(
  subData, 
  aes(x = Subtype, y = SER_map_Correlation1_tumor)
) +
  geom_boxplot()


```

Next, find a function from an R package that can create a "heat map" and
see if the features cluster together. Overlay your clustering with a
clinical feature. Does the clustering of image features seem to
correlate with the clinical feature?

```{r}
heatMatrix <- scale(subData[,-1])

row_ha <- rowAnnotation(Subtype = subData$Subtype)
Heatmap(
  heatMatrix, column_labels = rep("", ncol(heatMatrix)),
  right_annotation = row_ha
)

row_ha_her2 <- rowAnnotation(HER2 = ifelse(fullData$HER2 == 1, "HER2+", "HER2-"))
Heatmap(
  heatMatrix, column_labels = rep("", ncol(heatMatrix)),
  right_annotation = row_ha_her2
)

row_ha_surv <- rowAnnotation(Survival = fullData$survDays)
Heatmap(
  heatMatrix, column_labels = rep("", ncol(heatMatrix)),
  right_annotation = row_ha_surv
)

```

Heat maps can be a useful tool for finding broad patterns in data. In
the above example, there may be some clustering of textural features,
but the clustering does not correspond to any clinical outcomes.
Clustering does not provide rigorous statistical evidence of
associations or lack thereof, but it can be a useful exploratory step.

## Task 3: Preliminary Models

Consider HER2 positivity as a binary outcome. Fit a logistic regression
model using the covariates used above. Which variables are statistically
significant based on a significance level of 0.05?

```{r}
subData_her2 <- fullData %>% 
  select(HER2, all_of(tumorEnhanceTexture_vars))
her2Model <- glm(
  HER2 ~ ., data = subData_her2, family = binomial(link = "logit")
)
coefficients <- summary(her2Model)$coefficients
pVal <- coefficients[,"Pr(>|z|)"]
coefficients[pVal < 0.05, ]
```

This is an example of fitting a very basic statistical model. In general
you should not just blindly throw variables into a model and draw
statistical conclusions.

## Task 4: Survival Information

Fit a Kaplan-Meier survival curve to the time to death and time to
recurrence data. For each time to event outcome, fit a Kaplan-Meier
curve split by molecular subtype. Which subtypes have the most favorable
prognosis? Least favorable?

```{r}
deathFit <- survfit(Surv(survDays, dead) ~ 1, data = clinData)
autoplot(deathFit) +
  labs(title = "Kaplan-Meier Survival Curve, Time to Death",
       x = "Days", y = "Survival")

recurFit <- survfit(Surv(recurDays, recurrence) ~ 1, data = clinData)
autoplot(recurFit) +
  labs(title = "Kaplan-Meier Survival Curve, Time to First Recurrence",
       x = "Days", y = "Survival")

autoplot(update(deathFit, .~.+Subtype)) +
  labs(title = "Kaplan-Meier Survival Curve, Time to Death by Subtype",
       x = "Days", y = "Survival", col = "Subtype", fill = "Subtype")
autoplot(update(recurFit, .~.+Subtype)) +
  labs(title = "Kaplan-Meier Survival Curve, Time to Recurrence by Subtype",
       x = "Days", y = "Survival", col = "Subtype", fill = "Subtype")
```
