---
title: "Clinical Data Summary and Missingness"
author: "Grant Carr"
date: "`r Sys.Date()`"
output:
  html_document:
    df_print: paged
---

## R Setup

Load packages that are necessary for running all code below.
```{r}
#' example code for installing packages for first time:
#' install.packages("tidyverse")
library(tidyverse)
library(ggplot2)
library(survival)
library(ggfortify)
library(devtools)

#' install_github("jokergoo/ComplexHeatmap")
library(ComplexHeatmap)
library(gtsummary)
library(visdat)
library(survminer)
```


## Reading Clinical Data


```{r}
clinData <- read.csv(
  "/Users/grantcarr/Documents/Michigan/BDSI2025/Data/clinicalData_clean.csv"
)
clinData <- clinData[,-1] # remove first column of row numbers since we have
# patient ID
```


Subset clinical data to "relevant" columns; may depend on your research 
question.
```{r}
clinData_subset <- clinData %>% 
  select(
    Days.to.MRI..From.the.Date.of.Diagnosis.,
    Date.of.Birth..Days.:Race.and.Ethnicity,
    ER:Tumor.Location,
    Surgery:Age.at.last.contact.in.EMR.f.u.days..from.the.date.of.diagnosis...last.time.patient.known.to.be.alive..unless.age.of.death.is.reported.in.such.case.the.age.of.death,
    Neoadjuvant.Chemotherapy:Adjuvant.Endocrine.Therapy.Medications,
    Neoadjuvant.Anti.Her2.Neu.Therapy:Received.Neoadjuvant.Therapy.or.Not
  )
colnames(clinData_subset) <- c(
  "Days to MRI", "Age at Diagnosis", "Menopause at Diagnosis", "Race/Ethnicity",
  "ER", "PR", "HER2", "Subtype", "Oncotype Score", "T Stage", "N Stage", 
  "M Stage", "T Grade", "N Grade", "M Grade", "Nottingham Grade", "Histology", 
  "Location", "Surgery", "Time to Surgery", "Surgery Type", 
  "Neoadjuvant Radiation", "Adjuvant Radiation", "Clin Resp", "Path Resp", 
  "Recurrence", "Time to Local Recurrence", "Time to Distant Recurrence",
  "Days to Death", "Days to Last Local Free Assessment", 
  "Days to Last Distant Free Assessment",
  "Days to Last Follow-Up", "Neo Chemo", "Adjuvant Chemo",
  "Neo Endocrine", "Adjuvant Endocrine", "Neo HER2", "Adjuvant HER2",
  "Any Neoadjuvant"
)
```

```{r}
vis_dat(clinData_subset)
```

Time variables like days to recurrence are treated as characters/categorical.
Categorical variables like subtype are treated as numeric variables since they are coded as integers.

```{r}
clinData_subset <- clinData_subset %>% 
  mutate(
    `Age at Diagnosis` = -1*`Age at Diagnosis`/365,
    `Menopause at Diagnosis` = case_when(
      `Menopause at Diagnosis` == 0 ~ "Pre",
      `Menopause at Diagnosis` == 1 ~ "Post",
      `Menopause at Diagnosis` == 2 ~ NA,
      .default = NA
    ) %>% as.factor(),
    Location = as.factor(Location),
    `Race/Ethnicity` = as.factor(
      ifelse(`Race/Ethnicity` %in% 1:8, `Race/Ethnicity`, NA)
    ),
    ER = as.factor(ER),
    PR = as.factor(PR),
    HER2 = as.factor(HER2),
    Subtype = as.factor(Subtype),
    `T Stage` = as.factor(`T Stage`),
    `N Stage` = as.factor(`N Stage`),
    `M Stage` = as.factor(`M Stage`),
    `T Grade` = as.factor(`T Grade`),
    `N Grade` = as.factor(`N Grade`),
    `M Grade` = as.factor(`M Grade`),
    `Nottingham Grade` = as.factor(`Nottingham Grade`),
    Histology = as.factor(Histology),
    Surgery = as.factor(Surgery),
    `Time to Surgery` = as.numeric(`Time to Surgery`),
    `Surgery Type` = as.factor(`Surgery Type`),
    `Neoadjuvant Radiation` = as.factor(`Neoadjuvant Radiation`),
    `Adjuvant Radiation` = as.factor(`Adjuvant Radiation`),
    `Time to Local Recurrence` = as.numeric(
      `Time to Local Recurrence`
    ),
    `Time to Distant Recurrence` = as.numeric(
      `Time to Distant Recurrence`
    ),
    `Days to Death` = as.numeric(`Days to Death`),
    `Days to Last Local Free Assessment` = as.numeric(
      `Days to Last Local Free Assessment`
    ),
    `Days to Last Distant Free Assessment` = as.numeric(
      `Days to Last Distant Free Assessment`
    ),
    `Days to Last Follow-Up` = as.numeric(
      `Days to Last Follow-Up`
    ),
    `Neo Chemo` = as.factor(`Neo Chemo`),
    `Adjuvant Chemo` = as.factor(`Adjuvant Chemo`),
    `Neo Endocrine` = as.factor(`Neo Endocrine`),
    `Adjuvant Endocrine` = as.factor(`Adjuvant Endocrine`),
    `Neo HER2` = as.factor(`Neo HER2`),
    `Adjuvant HER2` = as.factor(`Adjuvant HER2`),
    `Any Neoadjuvant` = as.factor(`Any Neoadjuvant`)
  )
```

```{r}
vis_dat(clinData_subset)
```

Create event outcomes and times
```{r}
clinData_subset <- clinData_subset %>% 
  mutate(
    survDays = case_when(
      is.na(`Days to Death`) ~ pmax(
        `Days to Last Local Free Assessment`,
        `Days to Last Distant Free Assessment`,
        `Days to Last Follow-Up`, na.rm = T
      ),
      .default = `Days to Death`
    ) %>% as.numeric(),
    Death = ifelse(
      is.na(`Days to Death`), 0, 1
    ),
    recurDays = pmin(
      `Time to Local Recurrence`, 
      `Time to Distant Recurrence`, na.rm = T
    ) %>% ifelse(
      is.na(.), 
      pmax(
        `Days to Last Local Free Assessment`,
        `Days to Last Distant Free Assessment`, na.rm = T
      ), .
    ),
    recur = Recurrence,
    recurFreeSurv = pmin(
      recurDays, 
      survDays, na.rm = T
    ),
    recurFreeEvent = ifelse(
      Death == 1 | recur == 1, 1, 0
    )
  )

clinData_subset <- clinData_subset %>% mutate(
  recurrence_1year = ifelse(
    recur == 1 & recurDays <= 365, 1, 0
  ),
  recurrence_2year = ifelse(
    recur == 1 & recurDays <= 365*2, 1, 0
  ),
  recurrence_3year = ifelse(
    recur == 1 & recurDays <= 365*3, 1, 0
  ),
  recurrence_4year = ifelse(
    recur == 1 & recurDays <= 365*4, 1, 0
  ),
  recurrence_5year = ifelse(
    recur == 1 & recurDays <= 365*5, 1, 0
  )
)
```




```{r}
vis_miss(clinData_subset)
```

```{r}
table(
  clinData_subset[ , c("Clin Resp", "Path Resp")], 
  useNA = "always"
)
table(
  clinData_subset[ , c("Any Neoadjuvant", "Path Resp")], 
  useNA = "always"
)
table(
  clinData_subset[ , c("Any Neoadjuvant", "Clin Resp")], 
  useNA = "always"
)
```

Only clinical/pathological response if there is a treatment to respond to. 
When "Any Neoadjuvant" is 1 (yes), we have response measured. When it's 2 (no),
clinical/pathological response are NA.

```{r}
table(clinData_subset[ , c("recurrence_1year")])
table(clinData_subset[ , c("recurrence_2year")])
table(clinData_subset[ , c("recurrence_3year")])
table(clinData_subset[ , c("recurrence_4year")])
table(clinData_subset[ , c("recurrence_5year")])
```

```{r}
recurFreeKM <- survfit(Surv(recurFreeSurv, recurFreeEvent) ~ 1, data = clinData_subset)
recurKM <- survfit(Surv(recurDays, recur) ~ 1, data = clinData_subset)

ggsurvplot_combine(
  fit = list(recurKM, recurFreeKM), 
  data = clinData_subset,
  ylim = c(0.75, 1), 
  legend.labs = c("Recurrence", "Recurrence-Free Survival")
)

```


Table 1

```{r}
tbl_summary(
  clinData_subset
)

```

```{r}
tbl_summary(
  clinData_subset, by = "Subtype", 
  label = list(`Age at Diagnosis` = "Age (Years)"),
  statistic = list(
    all_continuous() ~ "{mean} ({sd})", 
    all_categorical() ~ "{n}"
  ), 
  missing_text = "Missing"
) %>% 
  add_p() %>% bold_p(t = 0.01) %>% 
  modify_header(label = "**Variable**")

show_header_names(
  tbl_summary(
    clinData_subset, by = "Subtype", 
  label = list(`Age at Diagnosis` = "Age (Years)"),
  statistic = list(
    all_continuous() ~ "{mean} ({sd})", 
    all_categorical() ~ "{n}"
  ), 
  missing_text = "Missing"
  )
)
```





